# Supplementary materials {#supplement}

## Beginner R

- Basic R functionality
- Reading in data
- Tidy R philosophy
- Manipulating data with [core `tidyr` functions](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

### Basic R functionality

- Variables
- Functions
- Operators

### Variables

The way R stores your information will determine the kinds of functions/operators you can use

- Data: dataframes
- Values: lists, vectors, matrices, etc.

```{r variables}
# Variables can be numbers, strings, etc.
some_val <- 200
other_val <- TRUE

# Variables can also be lists of numbers, strings, etc.
a_list <- c("a", "b", "c")

# We can also make dataframes (which are essentially special lists)
a_dataframe <- tibble(column_1 = 1:3,
                      column_2 = a_list)
```

### Functions

Functions take a certain number and certain types of "arguments"

- `base` R functions: part of downloading R
- Packages: need to be installed, then loaded
  - `install.packages(package)` once (or to update package)
  - `library(package)` every R session

Use Help window or `?package` to check argument names, types, and defaults

- Named arguments without values are required (and user-defined)
- Named arguments with values show defaults
- Elipses mean that you can add other optional arguments

```{r arguments, eval = FALSE}
# file: required
# header = TRUE: default
# ...: other potential arguments, like stringsAsFactors = FALSE
read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)
```

### `base` R

```{r base_functions_1}
# base R function
# Count the number of rows in this dataframe
nrow(a_dataframe)
```

```{r base_functions_2, eval = FALSE}
# If I give nrow() too many arguments, it will throw an error
nrow(a_dataframe, a_list)
```

```{r base_functions_3}
# If I give it the wrong kind of argument, it will just return NULL
# Some functions won't run at all with wrong kind of argument
nrow(a_list)
```

### Packages {#scholarex}

```{r pkg_functions}
# scholar package 
library(scholar)

# get_publications function
# Pull publications from Google Scholar for Marie Curie
get_publications("EmD_lTEAAAAJ&EmD_lTEAAAAJ&") %>% 
  dplyr::filter(cites > 30) %>% 
  distinct(title, .keep_all = TRUE) %>%
  select(author, title) %>%
  head(2) %>%
  kable()
```

### Operators

- Relational: >, <, ==, !=, <=, >=
  - `is.na()`, `exists()`, etc. will return TRUE/FALSE values
  - `grep()`, `filter()`, `str_detect()`, etc. use TRUE/FALSE values
- Logical: !, &, &&, |, ||
- Arithmetic

### Reading in data

General parameters for csv files

```{r csv, eval = FALSE}
read.csv("file_name.csv", 
         header = TRUE, 
         stringsAsFactors = FALSE, 
         check.names = FALSE, 
         na.strings = "")
```

Avoid special characters (including spaces) in file names, directories, and column headers!

- `readxl` package for Excel spreadsheets
- APIs for direct access to online data
  - `qualtRics` package for Qualtrics data
  - `ggmap` package for Google services (geolocation data)
- `read_table()` from `readr` package for text files

### Tidy R philosophy

- One variable per column
- One observation per row

### Manipulating data with `tidyr`

- `%>%`: pass the results of one function on to another
- `select()`: choose columns by name
- `mutate()`: add/change columns
- `filter()`: filter for (or out) rows
- `group_by()` and `summarise()`: perform operations on groups of data
- ~~`gather()` and `spread()`~~ `pivot_longer()` and `pivot_wider()`: condense multiple columns into one or the inverse
- `separate()` and `unite()`: split a column into multiple or the inverse

### Other helpful `tidyr` functions

- `slice()`: choose a row
- `pull()`: choose a column
- Helper functions for `select()` (e.g., `contains()`)
- `join` family of functions: combine datasets based on a shared unique identifier
- `union()`: combine datasets by rows (column names must be the same)
- `replace_na()`/`drop_na()`: alter/remove rows with NA values 

### Helpful `base` R functions

- `rbind()` and `cbind()`: add rows/columns
- `nrow()` and `ncol()`: count rows/columns
- `unique()`: pull unique values
- Indexing with `var$column` and `var[row, column]`
- `which()` with column/row indexing

### Other packages and functions

- `tibble` package for dataframes with `tibble()`
- `kableExtra()` for `kable()` tables
- `ggplot2` package for graphs (cheat sheet [here](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf))
- `factor()` for ordering text labels in graphs
- Use `na.rm = TRUE` argument (e.g., in `mean()`) to remove NA values from calculations

## Intermediate R

- Dealing with free-response text
- Dynamic variable creation and reference
- Helpful functions
- Other tips

### Free-response text

- Regular expressions (cheat sheet [here](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf))
- Pattern matching
  - Return position/value of elements that match a pattern: `grep()`, `agrep()`
  - Return TRUE/FALSE: `grepl()`, `agrepl()`, `str_detect()`
  - Change a pattern: `sub()`, `gsub()`, `replace()`
  - Return position of pattern for all elements: `regexpr()`, `gregexpr()`, `regexec()`
- Use `perl = TRUE` argument to handle especially complex patterns

```{r text_1}
# List of elements
fruit <- c("apple", "banana", "pear", "pinapple")

# grep position
grep(pattern = "le", x = fruit)

# grep value
grep(pattern = "le", x = fruit, value = TRUE)
```

```{r text_2}
# agrep (can also specify value = TRUE)
agrep(pattern = "le", x = fruit, max.distance = 0.1)

# regexpr
# match.length attribute gives starting position of match
# index.type attribute gives length of matched text
regexpr(pattern = "le", text = fruit)
```

### Dynamic variables

```{r assign_1}
# Variables
vals <- rep(1:3, 3)
name <- "assign_example"

# Assign values to variable name
assign(name, vals)

# Use the variable as usual
assign_example
```

```{r assign_2}
# You can also get the new variable from the name
get(name)

# You can add to this variable dynamically as well
assign(name, c(get(name), 4:6))

# New output
assign_example
```

### Helpful functions

- `apply()`, `lapply()`, `sapply()`, `tapply()`
- `source()` variables from R scripts
- `%notin%` and `%in%` (compared to != and ==)

```{r notin, eval = FALSE}
# Example using apply(): go across columns of dataset 
  # and substitute characters
language <- apply(language, 2, 
                  function(x) gsub("\\\\", "", x, fixed = TRUE))

# Source other sripts
source("data_cleaning.R", local = TRUE)

# Helper function
"%notin%" <- Negate("%in%")

# Example from processing pipeline for Qualtrics data
unusable <- c("0","00","107")
dat %>% dplyr::filter(Progress==100 & ID %notin% unusable)
```

### Other tips
- Store `ggplot2` parameters in a `list()`
- [`list()` vs. `c()`](https://stackoverflow.com/questions/8594814/what-are-the-differences-between-vector-and-list-data-types-in-r)

## Practice materials

You can find practice materials for R Markdown and R [here](https://github.com/hollzzar/markdown-tutorial). You'll want to clone this repository or download the following documents/folders:

- brownbag_activity.Rmd
- data_cleaning.R
- apa.csl
- /data