--- 
title: "R Markdown for Psychology Graduate Students"
author: "Holly Zaharchuk"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: packages.bib
biblio-style: apalike
link-citations: yes
github-repo: hollzzar/rmarkdown-guide
description: "Guide to using R Markdown"
always_allow_html: true
---

```{r setup, echo=FALSE}
## R setup ##

chooseCRANmirror(graphics = FALSE, ind = 1)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE, eval = TRUE)

## Load packages ##

# Package list
pkg_list <- c("plyr","tidyverse", "data.table", "ggplot2", "kableExtra")

# Load packages
pacman::p_load(pkg_list, character.only = TRUE)
```

# Welcome

I designed this guide to be a resource for psychology graduate students looking to streamline their research pipelines. With R Markdown, you can load, clean, manipulate, analyze, and present your data in one environment. This guide focuses on the presentation piece, with information on creating slides, posters, manuscripts, CVs, and reports in several formats, including HTML, PDF, and Microsoft Word.

## Background

This guide assumes a basic level of familiarity with [R](https://www.r-project.org/) and [RStudio](https://rstudio.com/products/rstudio/download/). If you don't know how to use either of these, there are several [beginner tutorials](https://education.rstudio.com/learn/beginner/) you should check out first. I have also created primers for Beginner and Intermediate R in Chapter \@ref(supplement).

## Getting started

What is R Markdown?

[Markdown](https://en.wikipedia.org/wiki/Markdown) is a specific [markup language](https://en.wikipedia.org/wiki/Markup_language) with plain text-formatting syntax. [R Markdown](https://cran.r-project.org/web/packages/stationery/vignettes/Rmarkdown.pdf) is a kind of markdown document

R Markdown and R are not the same thing! R Markdown combines R code (or code from other programming languages) and markdown in the RStudio integrated development environment (IDE). This allows you to embed code and text in the same document.

You can install the **R Markdown** package from CRAN or GitHub:

```{r eval=FALSE}
install.packages("rmarkdown")
# or the development version
# devtools::install_github("rstudio/rmarkdown")
```